\bf Visualising How Non-linear Dimension Reduction Warps Your Data

### 2024-03-11 paper.qmd
Added PaCMAP layout for handwritten data with preprocessing steps

@@

```{r}
#| echo: false
#| fig-cap: "Goodness of fit statistics from UMAP applied to training S-curve dataset. What is the effective number of bins in each NLDR technique to create a 2D model? The MSE plot have a steep slope at the beginning, indicating that a smaller number of bins causes a larger amount of error. Then, the slope gradually declines or level off, indicating that a higher number of bins generates a smaller error. Using the elbow method, when the total number of bins is set to 144, the slope of the Mean Squared Error (MSE) plot experiences a sudden and noticeable change, resembling an elbow-like shape. This point indicates that adding less bins does not enough to capture the data structure."
#| label: fig-diagnosticpltMNIST
##| out-width: 100%
#| fig-pos: H

@@

## Decide by looking at MSE plot
num_bins_x_mnist <- 6
num_bins_y_mnist <- 13
hex_size_mnist <- 0.14

@@

## To obtain the standardise counts within hexbins
counts_df <- as.data.frame(do.call(cbind, hb_obj_mnist$std_cts))
df_bin_centroids <- extract_hexbin_centroids(centroids_df = all_centroids_df, 
                                             counts_df = counts_df)
ggplot(data = hex_grid, aes(x = x, y = y)) + 
  geom_polygon(fill = "white", color = "black", aes(group = hex_poly_id)) +
  geom_point(data = all_centroids_df, aes(x = c_x, y = c_y), color = "red") +
  coord_fixed()

@@

@@

distance_plot <- plot_dist(distance_mnist) +
  ylab(expression(d^{(2)})) +
  theme(axis.text = element_text(size = 5),
        axis.title = element_text(size = 12))

@@

ggplot() +
geom_trimesh(data = df_bin_centroids_mnist, mapping = aes(x = c_x, y = c_y))

@@

error_df <- error_df |>
  mutate(type = if_else(total == 0, "no error",
                        if_else(total <= 80, "error 1-50",
                                if_else(total <= 100, "error 50-100",
                                        if_else(total <= 200, "error 100-200",
                                                if_else(total <= 300, "error 200-300",
                                                        if_else(total <= 400, "error 300-400",
                                                                if_else(total <= 500, "error 400-500", "error greter than 500"))))))))

plot_list2_mnist_error <- error_df |>
  mutate(type = factor(type , levels = c("no error", "error 1-50", "error 50-100", "error 100-200", "error 200-300", "error 300-400", "error 400-500", "error greter than 500"))) |>
    ggplot(aes(x = PaCMAP1,
               y = PaCMAP2, color = type))+
    geom_point(alpha=0.5, size = 0.1) +
    coord_equal() +
    theme(plot.title = element_text(hjust = 0.5, size = 18, face = "bold")) + #ggtitle("(a)") +
  theme_linedraw() +
    theme(plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
              axis.title.x = element_blank(), axis.title.y = element_blank(),
              axis.text.x = element_blank(), axis.ticks.x = element_blank(),
              axis.text.y = element_blank(), axis.ticks.y = element_blank(),
              panel.grid.major = element_blank(), panel.grid.minor = element_blank(), #change legend key width
        legend.title = element_text(size=5), #change legend title font size
        legend.text = element_text(size=4),
         legend.key.height = unit(0.25, 'cm'),
         legend.key.width = unit(0.25, 'cm')) +
  scale_color_manual(values=c("#fee0d2", "#fcbba1",
                                         "#fc9272", "#fb6a4a", "#ef3b2c",
                                         "#cb181d", "#a50f15", "#99000d"))

@@

```{r}
#| echo: false
#| fig-cap: "(a) 2D layout from UMAP (n_neighbors = 30, min_dist = 0.3) applied for the PBMC3k dataset. Is this a best representation of the original data?, (b)  Model in the 2D space with UMAP (<https://youtu.be/CNXjWZMyh_M>)"
#| label: fig-nldervisMNISTPACMAP
#| out-width: 100%
#| fig-pos: h

>>>>

#### Preprocessing

- Scale the pixel values to the range [0,1]

- Reshape to have a 2-dimensional array

-  reshaping pixel data is a fundamental preprocessing step in image analysis tasks, ensuring that the data is in a suitable format for further analysis, modeling, and interpretation. It helps preserve important spatial information, ensures compatibility with machine learning models, and facilitates feature extraction and visualization.

-  Reshaping the pixel data allows us to preserve this spatial information, ensuring that the relationships between neighboring pixels are maintained.

- Many machine learning algorithms, including neural networks, expect input data to be in a specific format. For image data, this typically means a two-dimensional array where each row represents a single sample (image), and each column represents a feature (pixel value). Reshaping ensures that all samples have consistent dimensions, making them suitable for feeding into these algorithms.

- Reshaping the pixel data allows for easier extraction of relevant features from the images. Once reshaped, individual pixels can be treated as features, making it easier to apply techniques like dimensionality reduction (e.g., PCA) or feature engineering to extract meaningful information from the images.

- PaCMAP parameters: n_components=2, n_neighbors=10, init="random", MN_ratio=0.9, FP_ratio=2.0

@@

```{r}
#| echo: false
#| fig-cap: "Goodness of fit statistics from PaCMAP applied to training MNIST dataset. What is the effective number of bins in each NLDR technique to create a 2D model? The MSE plot have a steep slope at the beginning, indicating that a smaller number of bins causes a larger amount of error. Then, the slope gradually declines or level off, indicating that a higher number of bins generates a smaller error. Using the elbow method, when the total number of bins is set to 80, the slope of the Mean Squared Error (MSE) plot experiences a sudden and noticeable change, resembling an elbow-like shape. This point indicates that adding less bins does not enough to capture the data structure."
#| label: fig-diagnosticpltMNIST
##| out-width: 100%
#| fig-pos: H

@@

## Decide by looking at MSE plot
num_bins_x_mnist <- 8
num_bins_y_mnist <- 10
hex_size_mnist <- 0.1

@@

## To obtain the standardise counts within hexbins
counts_df <- as.data.frame(do.call(cbind, hb_obj_mnist$std_cts))
df_bin_centroids <- extract_hexbin_centroids(centroids_df = all_centroids_df, 
                                             counts_df = counts_df)
# ggplot(data = hex_grid, aes(x = x, y = y)) + 
#   geom_polygon(fill = "white", color = "black", aes(group = hex_poly_id)) +
#   geom_point(data = all_centroids_df, aes(x = c_x, y = c_y), color = "red") +
#   coord_fixed()

@@

# ggplot(data = hex_grid_with_counts, aes(x = x, y = y)) +
#   geom_polygon(color = "black", aes(group = hex_poly_id, fill = std_counts)) +
#   geom_text(data = all_centroids_df, aes(x = c_x, y = c_y, label = hexID)) +
#   scale_fill_viridis_c(direction = -1, na.value = "#ffffff") +
#   coord_fixed()

@@

# distance_plot <- plot_dist(distance_mnist) +
#   ylab(expression(d^{(2)})) +
#   theme(axis.text = element_text(size = 5),
#         axis.title = element_text(size = 12))
# 
# distance_plot

@@

# ggplot() +
# geom_trimesh(data = df_bin_centroids_mnist, mapping = aes(x = c_x, y = c_y))

@@

error_df <- error_df |>
  mutate(type = if_else(total == 0, "no error",
                        if_else(total <= 2, "error 0-2",
                                if_else(total <= 4, "error 2-4",
                                        if_else(total <= 6, "error 4-6",
                                                if_else(total <= 8, "error 6-8",
                                                        if_else(total <= 12, "error 8-10",
                                                                if_else(total <= 14, "error 10-12", "error greter than 12"))))))))

plot_list2_mnist_error <- error_df |>
  mutate(type = factor(type , levels = c("no error", "error 0-2", "error 2-4", "error 4-6", "error 6-8", "error 8-10", "error 10-12", "error greter than 12"))) |>
    ggplot(aes(x = PaCMAP1,
               y = PaCMAP2, color = type))+
    geom_point(alpha=0.5, size = 0.1) +
    coord_equal() +
    theme(plot.title = element_text(hjust = 0.5, size = 18, face = "bold")) + #ggtitle("(a)") +
  theme_linedraw() +
    theme(plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
              axis.title.x = element_blank(), axis.title.y = element_blank(),
              axis.text.x = element_blank(), axis.ticks.x = element_blank(),
              axis.text.y = element_blank(), axis.ticks.y = element_blank(),
              panel.grid.major = element_blank(), panel.grid.minor = element_blank(), #change legend key width
        legend.title = element_text(size=5), #change legend title font size
        legend.text = element_text(size=4),
         legend.key.height = unit(0.25, 'cm'),
         legend.key.width = unit(0.25, 'cm')) +
  scale_color_manual(values=c("#fee0d2", "#fcbba1",
                                         "#fc9272", "#fb6a4a", "#ef3b2c",
                                         "#cb181d", "#a50f15", "#99000d"))

@@

```{r}
#| echo: false
#| fig-cap: "(a) 2D layout from UMAP (n_components=2, n_neighbors=10, init=random, MN_ratio=0.9, FP_ratio=2.0) applied for the digit 1 of the MNIST dataset. Is this a best representation of the original data?, (b)  Model in the 2D space with PaCMAP (<https://youtu.be/CNXjWZMyh_M>)"
#| label: fig-nldervisMNISTPACMAP
#| out-width: 100%
#| fig-pos: h

### 2024-03-11 paper.tex
Added PaCMAP layout for handwritten data with preprocessing steps

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, frame hidden, enhanced, boxrule=0pt, breakable, borderline west={3pt}{0pt}{shadecolor}, interior hidden]}{\end{tcolorbox}}\fi

@@

@@

\caption{\label{fig-diagnosticpltMNIST}Goodness of fit statistics from
UMAP applied to training S-curve dataset. What is the effective number
of bins in each NLDR technique to create a 2D model? The MSE plot have a
steep slope at the beginning, indicating that a smaller number of bins
causes a larger amount of error. Then, the slope gradually declines or
level off, indicating that a higher number of bins generates a smaller
error. Using the elbow method, when the total number of bins is set to
144, the slope of the Mean Squared Error (MSE) plot experiences a sudden
and noticeable change, resembling an elbow-like shape. This point
indicates that adding less bins does not enough to capture the data
structure.}

@@

\includegraphics{paper_files/figure-pdf/unnamed-chunk-73-1.pdf}

\includegraphics{paper_files/figure-pdf/unnamed-chunk-73-2.pdf}

\includegraphics{paper_files/figure-pdf/unnamed-chunk-73-3.pdf}

\includegraphics{paper_files/figure-pdf/unnamed-chunk-73-4.pdf}

@@

\caption{\label{fig-nldervisMNISTPACMAP}(a) 2D layout from UMAP
(n\_neighbors = 30, min\_dist = 0.3) applied for the PBMC3k dataset. Is
this a best representation of the original data?, (b) Model in the 2D
space with UMAP (\url{https://youtu.be/CNXjWZMyh_M})}

>>>>

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, interior hidden, breakable, sharp corners, enhanced, borderline west={3pt}{0pt}{shadecolor}, frame hidden]}{\end{tcolorbox}}\fi

@@

\hypertarget{preprocessing}{%
\subsubsection{Preprocessing}\label{preprocessing}}

\begin{itemize}
\item
  Scale the pixel values to the range {[}0,1{]}
\item
  Reshape to have a 2-dimensional array
\item
  reshaping pixel data is a fundamental preprocessing step in image
  analysis tasks, ensuring that the data is in a suitable format for
  further analysis, modeling, and interpretation. It helps preserve
  important spatial information, ensures compatibility with machine
  learning models, and facilitates feature extraction and visualization.
\item
  Reshaping the pixel data allows us to preserve this spatial
  information, ensuring that the relationships between neighboring
  pixels are maintained.
\item
  Many machine learning algorithms, including neural networks, expect
  input data to be in a specific format. For image data, this typically
  means a two-dimensional array where each row represents a single
  sample (image), and each column represents a feature (pixel value).
  Reshaping ensures that all samples have consistent dimensions, making
  them suitable for feeding into these algorithms.
\item
  Reshaping the pixel data allows for easier extraction of relevant
  features from the images. Once reshaped, individual pixels can be
  treated as features, making it easier to apply techniques like
  dimensionality reduction (e.g., PCA) or feature engineering to extract
  meaningful information from the images.
\item
  PaCMAP parameters: n\_components=2, n\_neighbors=10, init=``random'',
  MN\_ratio=0.9, FP\_ratio=2.0
\end{itemize}

@@

\caption{\label{fig-diagnosticpltMNIST}Goodness of fit statistics from
PaCMAP applied to training MNIST dataset. What is the effective number
of bins in each NLDR technique to create a 2D model? The MSE plot have a
steep slope at the beginning, indicating that a smaller number of bins
causes a larger amount of error. Then, the slope gradually declines or
level off, indicating that a higher number of bins generates a smaller
error. Using the elbow method, when the total number of bins is set to
80, the slope of the Mean Squared Error (MSE) plot experiences a sudden
and noticeable change, resembling an elbow-like shape. This point
indicates that adding less bins does not enough to capture the data
structure.}

@@

@@

\caption{\label{fig-nldervisMNISTPACMAP}(a) 2D layout from UMAP
(n\_components=2, n\_neighbors=10, init=random, MN\_ratio=0.9,
FP\_ratio=2.0) applied for the digit 1 of the MNIST dataset. Is this a
best representation of the original data?, (b) Model in the 2D space
with PaCMAP (\url{https://youtu.be/CNXjWZMyh_M})}
